Sample variance in weak lensing: how many simulations are required?

### 2016-01-09 Paper/paper.tex
addressed ZH2 comments

Constraining cosmology using weak gravitational lensing consists of
comparing a measured feature vector of dimension $N_b$ with its
simulated counterpart. An accurate estimate of the $N_b\times N_b$
feature covariance matrix $\bb{C}$ is essential to obtain accurate
parameter confidence intervals. When $\bb{C}$ is measured from a set
of simulations, an important question is how large this set should
be. To answer this question, we construct different ensembles of $N_r$
realizations of the shear field, using a common randomization
procedure that recycles the outputs from a smaller number $N_s\leq
N_r$ of independent ray-tracing $N$--body simulations.  We study
parameter confidence intervals as a function of ($N_s,N_r$) in the
range $1\leq N_s\leq 200$ and $1\leq N_r\leq 10^5$.  Previous work
\citep{DodelsonSchneider13} has shown that Gaussian noise in the
feature covariance lead, at quadratic order, to an $O(1/N_r)$
degradation of the parameter confidence intervals. Using a variety of
lensing features measured in our simulations, including shear-shear
power spectra and peak counts, we show that cubic and quartic
covariance fluctuations lead to additional $O(1/N_r^2)$ error
degradation that is not negligible when $N_r$ is only a factor of few
larger than $N_b$. We study the large $N_r$ limit, and find that a
single $N$--body simulation ($N_s=1$) can be repeatedly recycled to
produce as many as $N_r=10^5$ mutually independent shear maps. As a
result, a small number of simulations ($N_s=1$ or $2$) is sufficient
to forecast parameter confidence intervals at percent
accuracy.

@@

\subsection{Cosmological parameter inference}
%
Let $\bbh{d}$ be a single estimate for a feature of dimension $N_b$,
$\bb{d}(\bb{p})$ be the true value of this feature at a point $\bb{p}$
in parameter space (which has a dimension $N_p$) and $\bb{C}$ be the
$N_b\times N_b$ feature covariance matrix. For the purpose of this
work $\bb{p}$ is the triplet $(\Omega_m,w,\sigma_8)$ and $\bb{d}$ is
one of the features -- either a power spectrum or a peak count
histogram -- in Table \ref{featuretable}. Although existing emulators
can be used, in principle, to compute both $\bb{d}(\bb{p})$ and
$\bb{C}$, the latter is more difficult, and typically only the mean,
$\bb{d}(\bb{p})$, has been computed to date (refs.~\citep{coyote2,Nicaea}, but see
an exception by ref.~\citep{SchneiderKnoxCovariance}).
%ZH: I am missing the point here.  Why is covariance matrix ``more complicated'' ?
%    Even to estimate a mean, one has to ``generate a series of mock realizations''.
%    Unless you are willing to equate a single realization as the mean.   But then
%    by the same logic, you could estimate the covariance matrix from two realizations.
%    Please clarify what you mean...  maybe I am missing something?  Or you simply mean
%    that published emulators usually do not emulate the cov.matrix (even though they could)?
%    Note that there is a paper by Schneider and Knox, in which they do an interpolation of the
%    power spectrum covariance matrix.
%AP: The intended meaning is the latter: published emulators only predict the means. It is also true that getting the covariance accurate is more difficult than getting the means accurate. I rephrased the sentence.
%ZH2: OK, I rephrased it a bit above to be clearer.  I believe only Schneider et al. did ``C''.
Estimating $\bb{C}$ from simulations involves generating a series of
mock realizations $\bbh{d}_r$ with $r=1...N_r$ and computing the
sample covariance $\bbh{C}$,

@@

\begin{table*}
\begin{center}
\begin{tabular}{c|c|c||c|c}
\toprule
$\Omega_m$ &  $w$ & $\sigma_8$ & $(N_s,N_r)$ & Number of $\kappa$ ensembles \\ \hline \hline
\midrule
%
0.26 & $-$1 & 0.8 & (1 to 200,1024) & 16 \\
0.26 & $-$1 & 0.8 & (1,128000) & 1 \\
0.29 & $-$1 & 0.8 & (1,1024) & 1 \\
0.26 & $-$0.8 & 0.8 & (1,1024) & 1 \\
0.26 & $-$1 & 0.6 & (1,1024) & 1 \\
%
\bottomrule
\end{tabular}
\end{center}
\caption{Summary of the shear ensembles used in this work. $N_s$ and
  $N_r$ refer to the number of independent N-body simulations, and the
  number of pseudo-independent realizations created from these
  simulations, respectively.}
%ZH: So each of these have N_r=1000, yes?  You should state this here. And also,
%    Add the ensemble with N_s=1, N_r=10^5 to this table.
%    Also: point the reader to this table in the text, up front, when the 
%    ensembles are first discussed.
%AP: done
%ZH2: The first raw was a bit confusing, so I separated it into two rows. Also added words to the caption
%    (also, the N_r=10^5 one is the 17th ensemble, in addition to the 16 you listed here before, no? So 16 should have been 17?)
\label{simtable}
\end{table*}

@@

@@

mistakenly interpreted as a parameter bias.  We derived the scaling

@@

above by assuming that the noise in $\bbh{d}$ is Gaussian. We then
directly test these scaling relations using our simulations, in the
limits of both high and low $N_r$.

@@

@@

%ZH: reworded above.  OK?  Also, I am not sure if this point is new.
%    Didn't Hartlap (and others) make this point? If so, we should cite them.
%    Also, it would make the point more relevant if we could cite some 
%    papers (including our own) which neglected this correction in the past
%    and therefore under-forecasted the errors.
%AP: yes, Hartlap et al. reached a similar conclusion. I think we are saying something more though: even if the bias in the Psi estimator is corrected, if the same simulation set for Psi and C is used, parameter errors are still underestimated because of the explanation we give in section II-B. I don't think this point appears in Hartlap et al. 
%ZH2: OK I added this clarification above.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%% APPENDIX %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%ZH2: I went through the appendix, too, now, and made minor cosmetic changes.
%     
\section*{Appendix A: cubic and quartic covariance fluctuations}
\label{appendixA}

The goal of this appendix is to give a derivation of
eq. (\ref{quarticdegradation}). When the simulated feature vector
$\bbh{d}_r$ is drawn from a Gaussian distribution, the covariance
estimator $\bbh{C}$ follows the Wishart distribution, and its inverse
$\bbh{\Psi}$ follows the inverse Wishart distribution (see
Ref.~\citep{Taylor12} for analytical expressions for these probability
distributions). Computing expectation values of
eq. (\ref{estimatorcovariance}) over the inverse Wishart distribution
is not possible analytically, and a perturbative expansion is
necessary. Writing $\bbh{\Psi}=\bb{\Psi}+\delta\bbh{\Psi}$, we can
expand eq. (\ref{quarticdegradation}) 
%ZH2: you must mean eq.~(\ref{estimatorcovariance}) above, instead of eq.10 
in powers of $\delta\bbh{\Psi}$. The expectation value of each term in
this expansion can be calculated in terms of moments of the inverse
Wishart distribution. Ref.~\citep{MasumotoWishart} provides a general
framework to compute these moments, and give exact expressions for
moments up to quartic order. First, let us expand the inverse of the
Fisher matrix estimator (eq.~\ref{estimatorfisher}) in powers of
$\delta\bbh{\Psi}$. The $n$--th order of this expansion will be

>>>>

Constraining cosmology using weak gravitational lensing consists of
comparing a measured feature vector of dimension $N_b$ with its
simulated counterpart. An accurate estimate of the $N_b\times N_b$
feature covariance matrix $\bb{C}$ is essential to obtain accurate
parameter confidence intervals. When $\bb{C}$ is measured from a set
of simulations, an important question is how large this set should
be. To answer this question, we construct different ensembles of $N_r$
realizations of the shear field, using a common randomization
procedure that recycles the outputs from a smaller number $N_s\leq
N_r$ of independent ray-tracing $N$--body simulations.  We study
parameter confidence intervals as a function of ($N_s,N_r$) in the
range $1\leq N_s\leq 200$ and $1\leq N_r\lesssim 10^5$.  Previous work
\citep{DodelsonSchneider13} has shown that Gaussian noise in the
feature covariance lead, at quadratic order, to an $O(1/N_r)$
degradation of the parameter confidence intervals. Using a variety of
lensing features measured in our simulations, including shear-shear
power spectra and peak counts, we show that cubic and quartic
covariance fluctuations lead to additional $O(1/N_r^2)$ error
degradation that is not negligible when $N_r$ is only a factor of few
larger than $N_b$. We study the large $N_r$ limit, and find that a
single $N$--body simulation ($N_s=1$) can be repeatedly recycled to
produce as many as $N_r=10^5$ mutually independent shear maps. As a
result, a small number of simulations ($N_s=1$ or $2$) is sufficient
to forecast parameter confidence intervals at percent
accuracy.

@@

\subsection{Cosmological parameter inference}
%
Let $\bbh{d}$ be a single estimate for a feature of dimension $N_b$,
$\bb{d}(\bb{p})$ be the true value of this feature at a point $\bb{p}$
in parameter space (which has a dimension $N_p$) and $\bb{C}$ be the
$N_b\times N_b$ feature covariance matrix. For the purpose of this
work $\bb{p}$ is the triplet $(\Omega_m,w,\sigma_8)$ and $\bb{d}$ is
one of the features -- either a power spectrum or a peak count
histogram -- in Table \ref{featuretable}. Although existing emulators
can be used, in principle, to compute both $\bb{d}(\bb{p})$ and
$\bb{C}$, the latter is more difficult, and typically only the mean,
$\bb{d}(\bb{p})$, has been computed to date (refs.~\citep{coyote2,Nicaea}, but see
an exception by ref.~\citep{SchneiderKnoxCovariance}).
%ZH: I am missing the point here.  Why is covariance matrix ``more complicated'' ?
%    Even to estimate a mean, one has to ``generate a series of mock realizations''.
%    Unless you are willing to equate a single realization as the mean.   But then
%    by the same logic, you could estimate the covariance matrix from two realizations.
%    Please clarify what you mean...  maybe I am missing something?  Or you simply mean
%    that published emulators usually do not emulate the cov.matrix (even though they could)?
%    Note that there is a paper by Schneider and Knox, in which they do an interpolation of the
%    power spectrum covariance matrix.
%AP: The intended meaning is the latter: published emulators only predict the means. It is also true that getting the covariance accurate is more difficult than getting the means accurate. I rephrased the sentence.
%ZH2: OK, I rephrased it a bit above to be clearer.  I believe only Schneider et al. did ``C''.
%AP2: OK
Estimating $\bb{C}$ from simulations involves generating a series of
mock realizations $\bbh{d}_r$ with $r=1...N_r$ and computing the
sample covariance $\bbh{C}$,

@@

\begin{table*}
\begin{center}
\begin{tabular}{c|c|c||c|c}
\toprule
$\Omega_m$ &  $w$ & $\sigma_8$ & $(N_s,N_r)$ & Number of $\kappa$ ensembles \\ \hline \hline
\midrule
%
0.26 & $-$1 & 0.8 & (1 to 200,1024) & 16 \\
0.26 & $-$1 & 0.8 & (1,128000) & 1 \\
0.29 & $-$1 & 0.8 & (1,1024) & 1 \\
0.26 & $-$0.8 & 0.8 & (1,1024) & 1 \\
0.26 & $-$1 & 0.6 & (1,1024) & 1 \\
%
\bottomrule
\end{tabular}
\end{center}
\caption{Summary of the shear ensembles used in this work. $N_s$ and
  $N_r$ refer to the number of independent N-body simulations, and the
  number of pseudo-independent realizations created from these
  simulations, respectively.}
%ZH: So each of these have N_r=1000, yes?  You should state this here. And also,
%    Add the ensemble with N_s=1, N_r=10^5 to this table.
%    Also: point the reader to this table in the text, up front, when the 
%    ensembles are first discussed.
%AP: done
%ZH2: The first row was a bit confusing, so I separated it into two rows. Also added words to the caption
%    (also, the N_r=10^5 one is the 17th ensemble, in addition to the 16 you listed here before, no? So 16 should have been 17?)
%AP2: OK
\label{simtable}
\end{table*}

@@

%AP2: OK

@@

mistakenly interpreted as a parameter bias. We test scaling

@@

againts our simulations, in the limits of both high and low $N_r$.

@@

%AP2: Technically the 1/Nr scaling has been obtained using the inverse Wishart distribution; if the data is non Gaussian, the covariance is not Wishart, but maybe one can say that for an arbitrary, not too crazy, distribution the leading term is still 1/Nr and the coefficient is not too different from Nb-Np... This is indeed what we find: the error degradation is basically the same should we use power spectra or peak counts which have different PDFs... I removed the mention to the Gaussianity, which is not what we test. What we test is eq. 10 in the low Nr limit and the (possible) deviations from 1/Nr when Nr is very large (due to the maps being pseudo independent). 

@@

%ZH: reworded above.  OK?  Also, I am not sure if this point is new.
%    Didn't Hartlap (and others) make this point? If so, we should cite them.
%    Also, it would make the point more relevant if we could cite some 
%    papers (including our own) which neglected this correction in the past
%    and therefore under-forecasted the errors.
%AP: yes, Hartlap et al. reached a similar conclusion. I think we are saying something more though: even if the bias in the Psi estimator is corrected, if the same simulation set for Psi and C is used, parameter errors are still underestimated because of the explanation we give in section II-B. I don't think this point appears in Hartlap et al. 
%ZH2: OK I added this clarification above.  
%AP2: OK: Hartlap focuses on the Psi distribution, translating it into a parameter distribution using a naive Fisher analysis; what he does is in fact use the same simulation set for Psi and C, but this kind of discussion is absent from his paper.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%% APPENDIX %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%ZH2: I went through the appendix, too, now, and made minor cosmetic changes.
%AP2: OK     
\section*{Appendix A: cubic and quartic covariance fluctuations}
\label{appendixA}

The goal of this appendix is to give a derivation of
eq. (\ref{quarticdegradation}). When the simulated feature vector
$\bbh{d}_r$ is drawn from a Gaussian distribution, the covariance
estimator $\bbh{C}$ follows the Wishart distribution, and its inverse
$\bbh{\Psi}$ follows the inverse Wishart distribution (see
Ref.~\citep{Taylor12} for analytical expressions for these probability
distributions). Computing expectation values of
eq. (\ref{estimatorcovariance}) over the inverse Wishart distribution
is not possible analytically, and a perturbative expansion is
necessary. Writing $\bbh{\Psi}=\bb{\Psi}+\delta\bbh{\Psi}$, we can
expand eq.~(\ref{estimatorcovariance}) 
%ZH2: you must mean eq.~(\ref{estimatorcovariance}) above, instead of eq.10 
%AP2: yes
in powers of $\delta\bbh{\Psi}$. The expectation value of each term in
this expansion can be calculated in terms of moments of the inverse
Wishart distribution. Ref.~\citep{MasumotoWishart} provides a general
framework to compute these moments, and give exact expressions for
moments up to quartic order. First, let us expand the inverse of the
Fisher matrix estimator (eq.~\ref{estimatorfisher}) in powers of
$\delta\bbh{\Psi}$. The $n$--th order of this expansion will be
